---
title: "Getting Started with GOcontext"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Getting Started with GOcontext}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

## Overview

`GOcontext` provides tools for constructing **restricted Gene Ontology
(GO) graphs** prior to enrichment analysis.

In standard GO enrichment workflows, all GO terms within an ontology
(Biological Process, Molecular Function, or Cellular Component) are
tested simultaneously. However, for many biological questions, large
parts of the ontology are irrelevant to the experimental system under
study. Testing these terms unnecessarily increases the number of
hypotheses and therefore the multiple testing burden.

`GOcontext` addresses this by enabling users to construct
**context-specific GO subgraphs** before performing enrichment
analysis. The package:

- loads the GO structure from `GO.db`
- attaches organism-specific GO annotations from an `OrgDb`
- restricts the ontology graph to mapped terms or selected branches
- exports the resulting mappings for enrichment tools

By reducing the number of tested GO terms while preserving the
underlying directed acyclic graph (DAG) structure, enrichment analyses
can focus on biologically relevant parts of the ontology while
maintaining a transparent and reproducible GO universe.

The typical workflow is:

load_go() → attach_org() → filter_mapped() → subset_go() → enrichment

## Load packages

```{r Load packages}
suppressPackageStartupMessages({
    library(GOcontext)
    library(knitr)
    library(org.EcK12.eg.db)
})
```

## 1. Load a GO ontology

`GOcontext` loads GO directly from the Bioconductor package `GO.db`.
Only a single ontology (BP, MF, or CC) is loaded at a time.

```{r load_go}
go_bp <- GOcontext::load_go("BP")
```

The returned object contains:

- GO term metadata
- parent–child edges defining the GO DAG
- adjacency lists for graph traversal
- the GO.db version used to construct the graph

## 2. Attach organism annotations

GO enrichment requires mappings between GO terms and genes. These are
obtained from an organism annotation package (`OrgDb`).

```{r attach_org}
go_bp <- GOcontext::attach_org(
    go = go_bp,
    OrgDb = org.EcK12.eg.db
)
```

This attaches a GO-to-gene mapping table to the GO graph object.

## 3. Restrict the graph to mapped terms

Many GO terms have no gene annotations for a given organism. These
terms cannot contribute to enrichment results and can therefore be
removed before further analysis.

```{r filter_mapped}
go_bp_mapped <- GOcontext::filter_mapped(go_bp)
```

The returned object is a `GOSubgraph` containing only terms that have
at least one gene mapping.

This step often reduces the ontology size substantially.

## 4. Restrict the graph to selected branches

`GOcontext` allows restricting the ontology graph to specific regions.

Two modes are supported:

- **keep** – retain selected terms and all their descendants
- **exclude** – remove selected terms and their descendants

Example: restrict the graph to the descendants of a particular GO
term.

```{r subset_go}
go_sub <- GOcontext::subset_go(
    go = go_bp_mapped,
    ids = "GO:0006740",
    mode = "keep"
)
```

The returned `GOSubgraph` represents the restricted GO graph.

The object contains:

- retained GO terms
- removed GO terms
- induced parent–child edges
- updated adjacency lists
- the attached GO-to-gene mapping

## 5. Export mappings for enrichment

After restricting the graph, GO annotations can be exported in formats
commonly used by enrichment tools.

### TERM2GENE mapping

```{r as_term2gene}
term2gene <- GOcontext::as_term2gene(
    go        = go_sub,
    minGSSize = 10,
    maxGSSize = 500
)
head(term2gene)
```

### TERM2NAME mapping

```{r as_term2name}
term2name <- GOcontext::as_term2name(
    go        = go_sub,
    minGSSize = NULL,
    maxGSSize = NULL
)
head(term2name)
```

These tables can be supplied directly to
`clusterProfiler::enricher()`.

## 6. Run enrichment analysis

Example workflow with `clusterProfiler`:

```{r enrichment analysis, eval = FALSE}
library(clusterProfiler)

enrichment_results <- clusterProfiler::enricher(
    gene      = gene_list,
    TERM2GENE = term2gene,
    TERM2NAME = term2name
)
```

Because the GO graph was restricted beforehand, enrichment is
performed on a smaller and biologically focused hypothesis space.

## Interactive exploration

`GOcontext` also provides a console browser for exploring the GO
hierarchy interactively.

```{r browse go, eval = FALSE}
res <- GOcontext::browse_go(go_bp)
```
 

The browser allows navigation through the GO DAG and selection of
terms of interest. The selected terms can then be used as input for
`subset_go()`.

## Summary

`GOcontext` enables reproducible construction of restricted GO graphs
prior to enrichment analysis.

A typical workflow consists of:

1. Loading GO using `load_go()`
2. Attaching organism annotations with `attach_org()`
3. Restricting the ontology to mapped terms using `filter_mapped()`
4. Restricting the graph structure with `subset_go()`
5. Exporting GO mappings with `as_term2gene()` and `as_term2name()`
6. Running enrichment analysis

By explicitly defining the tested GO universe before enrichment,
`GOcontext` reduces the multiple testing burden while preserving the
ontology structure and ensuring transparent analysis workflows.

```{r session info}
sessionInfo()
```
