---
title: "Motivation"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Motivation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Motivation

The Gene Ontology (GO) is one of the most widely used resources for
functional enrichment analysis. Across its three ontologies —
Biological Process (BP), Molecular Function (MF), and Cellular
Component (CC) — thousands of GO terms are available as potential
hypotheses in enrichment testing.

In many workflows, the full ontology is tested by default. However,
not all GO terms are meaningful for a given organism or biological
question. As a result, enrichment analyses often include large numbers
of hypotheses that are biologically irrelevant.

For example, consider an RNA-seq experiment performed in *E. coli*.
The GO Biological Process ontology contains terms such as *heart
development*, *ossification*, *pregnancy*, or *neuron differentiation*.
These processes are biologically impossible in bacteria, yet they
remain part of the hypothesis space when the full ontology is tested.

In practice, analysts often deal with this implicitly. Common strategies
include filtering annotations, removing specific branches of the GO
graph, or restricting analyses to annotated terms. However, such steps
are frequently implemented through small ad hoc scripts and may not be
explicitly documented.

`GOcontext` does not introduce a fundamentally new statistical method.
Instead, it provides tools that make a common practice — restricting the
GO hypothesis space prior to enrichment analysis — easier and more
transparent.

Importantly, removing hypotheses *a priori* does **not** invalidate
statistical error control. Procedures such as Bonferroni (FWER control)
or Benjamini–Hochberg (FDR control) define their guarantees with
respect to the set of hypotheses that are actually tested. If
biologically irrelevant hypotheses are excluded before testing,
the statistical guarantees remain valid.

At the same time, reducing the hypothesis space decreases the multiple
testing burden and can therefore increase statistical power.

The following sections illustrate this effect analytically and through
simulation.

# Analytical effect of hypothesis space reduction

The effect of hypothesis restriction follows directly from the formulas
of multiple-testing correction procedures.

For Bonferroni correction (controlling the family-wise error rate), the
rejection threshold is

\[
p \le \frac{\alpha}{m},
\]

where \( m \) denotes the number of tested hypotheses.

If the hypothesis space is reduced *a priori* to

\[
m' = (1 - r)m
\]

where \( r \) is the proportion of excluded hypotheses, the new
threshold becomes

\[
p \le \frac{\alpha}{m'} = \frac{\alpha}{(1-r)m}.
\]

The ratio between the new and original thresholds is therefore

\[
\frac{\alpha/m'}{\alpha/m} = \frac{m}{m'} = \frac{1}{1-r}.
\]

Thus, excluding 20% of hypotheses (\( r = 0.2 \)) increases the
critical value by a factor of \( 1/(1-0.2) = 1.25 \), corresponding to a
25% relaxation.

The same scaling effect appears in the Benjamini–Hochberg (BH) false
discovery rate procedure. At rank \( k \), the BH rejection rule is

\[
p_{(k)} \le \frac{k}{m} q.
\]

After restricting the hypothesis space,

\[
p_{(k)} \le \frac{k}{m'} q = \frac{k}{(1-r)m} q.
\]

The Benjamini–Yekutieli (BY) procedure introduces an additional
harmonic correction factor, but the dominant dependence on \( 1/m \)
remains. In all cases, reducing the number of hypotheses produces a
deterministic relaxation of the rejection threshold.

# Critical value multiplier

The deterministic effect of hypothesis restriction can be summarized
as a **critical value multiplier**, defined as

\[
\text{multiplier} = \frac{m}{m'} = \frac{1}{1-r}.
\]

This multiplier depends only on the excluded proportion \( r \) and is
independent of the observed data.

```{r figure 1, echo = FALSE}
fig <- file.path("figures", "critical_value_multiplier.svg")
knitr::include_graphics(fig)
```

**Figure 1.** Critical value multiplier under a priori hypothesis
restriction. The multiplier represents the factor by which the
statistical rejection threshold increases when a proportion \( r \) of
hypotheses is removed before testing.

As the excluded proportion increases, the rejection boundary expands
nonlinearly. Even moderate reductions of the hypothesis space can lead
to noticeable relaxation of statistical thresholds.

# Empirical power gains under Monte Carlo simulation

To examine how the deterministic relaxation of critical values
translates into practical performance, we performed a Monte Carlo
simulation study.

Simulations were conducted separately for the three GO ontologies
(BP, MF, and CC), using hypothesis counts approximating their current
sizes. For each ontology, the number of true signals was fixed at

\[
s = 100.
\]

Test statistics were generated under a two-component mixture model:

- true signals were sampled from a normal distribution with mean shift 
(\( \mu = 3 \))
- null hypotheses were sampled from a standard normal distribution

Two-sided p-values were computed from the resulting z-statistics.

To simulate *a priori* hypothesis restriction, a proportion \( r \) of
null hypotheses was removed before testing, while the number of true
signals remained constant.

For each reduced hypothesis space

\[
m' = (1-r)m
\]

statistical power (true positive rate) was estimated across repeated
Monte Carlo replicates under three multiple testing procedures:

- Bonferroni (FWER control)
- Benjamini–Hochberg (BH; FDR control)
- Benjamini–Yekutieli (BY; FDR control under arbitrary dependence)

Importantly, the statistical procedure itself was unchanged. Only the
number of tested hypotheses was reduced before applying the correction.
Thus, any observed gain in power arises exclusively from the reduction
of multiplicity.

```{r figure 2, echo = FALSE}
fig <- file.path("figures", "power_gain_hypothesis_restriction_go.svg")
knitr::include_graphics(fig)
```

**Figure 2.** Empirical power gains from a priori hypothesis restriction
under Monte Carlo simulation. Only the number of tested hypotheses is
reduced; the testing procedure itself remains unchanged.

Across all procedures and ontology sizes, decreasing the hypothesis
space leads to systematic increases in statistical power. The magnitude
of the gain depends on the correction method and the total number of
tests but follows the deterministic scaling behavior derived above.

# Practical implications

These results illustrate a simple principle: reducing the number of
tested hypotheses can increase sensitivity without violating formal
error control, provided that the restriction is defined *before*
testing.

In the context of Gene Ontology analysis, many restrictions are both
obvious and biologically justified. Common examples include:

- removing GO terms without annotations for the organism of interest
- excluding ontology branches that are irrelevant for the system under study
- focusing analyses on specific biological regions of the GO graph

Such restrictions are frequently applied informally in enrichment
analyses. Making them explicit improves transparency and
reproducibility while also reducing the multiple testing burden.

`GOcontext` provides simple tools to construct restricted GO graphs and
export the resulting annotation mappings for enrichment workflows.

```{r session info}
sessionInfo()
```
